# File of approximation of the true causal mean by Monte Carlo integration

### Set the sample size and number of time points
```{r}
n = 1e7; time = 5
```

### Set the true parameters for the zero inflated poisson model
### First line is the bernoulli component, second is the poisson component
```{r}
z_i = -12; z_c = -4; z_z1 = 0.4; z_y1 = -0.4; z_x = 0.8
z_ip = -5.5; z_cp = -1; z_z1p = 0.01; z_y1p = 0.2; z_xp = 0.01
```

### Set the true parameters for the study outcome model
### s_y = 0.88 corresponds to an R^2 of 0.7
```{r}
y_i = 14; y_c = -2; y_z1 = -0.02; y_y1 = 0.1; y_x = 0.05; y_z = 2*y_z1; s_y = 0.88
```

### The expit function required to generate logit link bernoulli data
```{r}
expit = function(x){1/(1+exp(-x))}
```

### Data generation
```{r}
set.seed(1)
x = rnorm(n,mean=50,sd=1)
z0 = ifelse(rbinom(n,1,expit(cbind(1,x)%*%c(z_ip,z_xp)))>0,0,
	rpois(n,lambda=cbind(1,x)%*%c(z_i,z_x)))
z1 = ifelse(rbinom(n,1,expit(cbind(1,1,z0,x)%*%c(z_ip,z_cp,z_z1p,z_xp)))>0,0,
	rpois(n,lambda=cbind(1,1,z0,x)%*%c(z_i,z_c,z_z1,z_x)))
y1 = rnorm(n,mean=cbind(1,1,z0,x,z1)%*%c(y_i,y_c,y_z1,y_x,y_z),sd=s_y)

tcm = cbind(1,1,z1,y1,z0,0,x)
colnames(tcm) = c("Int","C","Zij","Yij","Zij1","Yij1","X")
nc = ncol(tcm)

for(i in 1:(time-1)){
	inds = (1+n*(i-1)):(n*i)
	zn = ifelse(rbinom(n,1,expit(cbind(1,1,tcm[inds,c(3:4,nc)])%*%c(z_ip,z_cp,z_z1p,z_y1p,z_xp)))>0,0,
		rpois(n,lambda=cbind(1,1,tcm[inds,c(3:4,nc)])%*%c(z_i,z_c,z_z1,z_y1,z_x)))
	yn = rnorm(n,mean=cbind(1,1,tcm[inds,c(3:4,nc)],zn)%*%c(y_i,y_c,y_z1,y_y1,y_x,y_z),sd=s_y)
	tcm = rbind(tcm,cbind(1,1,zn,yn,tcm[inds,3:4],x))
}
```

### True causal mean
```{r}
print(mean(tcm[(n*(time-1)):nrow(tcm),4]))
```